Deep learning for quality control of surface physiographic fields using satellite Earth observations

### 2023-10-17 docs/submission/response_to_reviewer_comments/merged/response_reviewer_1.tex

\documentclass[10pt]{article}
\usepackage{vmargin}
\setpapersize{USletter}
\usepackage{ragged2e}
\usepackage{hyperref}

\author{Tom Kimpson \\ 
	\and Margarita Choulga\\ 
 \and Matthew Chantry\\
\and Gianpaolo Balsamo\\  
 \and Souhail Boussetta\\
  \and Peter Dueben\\
   \and Tim Palmer\\
}
\title{\normalsize Response to Reviewer  Comments
  concerning HESS submission egusphere-2022-1177}
\begin{document}
	\maketitle
\noindent Author response to referee reports for the paper egusphere-2022-1177,
entitled \textit{``Deep Learning for Verification of Earth-System Parametrisation of Water Bodies"}. \newline

\noindent We thank the referees for their reading, helpful criticism and suggestions towards the improvement of the manuscript. \newline

@@

\noindent We have addressed each point in turn below and the manuscript has been updated accordingly \newline

\noindent  We hope that this satisfies the request for changes necessary to proceed with the publication of the updated manuscript. \newline

\newpage 
\section{Reviewer 1}

\subsection*{Essential comments}

@@

			This is an good point. For this work we are satisfied to consider the combined affect of monthly maps and salt lakes, since many of the locations we specifically highlight in the manuscript are saline lakes with large expected time variability in the surface water. Further work in this area is currently ongoing and we defer a more in depth, global study of saline lakes and monthly maps for the future.

			\item \textbf{Essential comment 12: What are shadows on Fig. 9-10? Please explain.}

		 Previously these shadows were confidence intervals. In the updated manuscript, with multiple VESPER trainings these shadows are the $\pm 1 \sigma$ bounds. We have specified this in the Figure captions.

	\subsection*{Other comments}
	All typos and editorial comments have been corrected in the updated manuscript.

@@

\end{enumerate}

\newpage 
\section{Reviewer 2}

\subsection*{Comments}
The reviewer suggests a major overhaul of the structure of the manuscript. We agree that this is a very good suggestion and the text has been completely restructured as recommend. We now present the construction of VESPER much more thoroughly, detail the various input fields, and specify the differences between the various VESPER generations.  Only then do we then go on to deploy VESPER on lake fields and discuss the results. Where possible we have made an effort to be more concise.

\begin{itemize}
	\item \textbf{l.54} The terminology re parameters, model and physiography has been updated throughout the text. The tables have also been updated to describe the choice of variables, and the different VESPER models  (see e.g. Tables 1-3 in update manuscript). 
	\item \textbf{l. 123 ERA5} All required information has been added to the text. 
	\item \textbf{l. 133 MODIS} All required information has been added to the text. 
	\item \textbf{l. 146 and Figure 3} By 4km resolution, we were referring to the resolution at the equator. This has now been updated in the text. Re the number of points at high latitudes, this is a natural consequence of the MODIS orbit, see e.g. animation at \url{https://svs.gsfc.nasa.gov/3348}
	\item \textbf{l. 184} This change has been made and a new table (Table 3) added which specified each VESPER configuration.
	\item \textbf{Figure 4 and related text}

	The prediction error is now defined at the end of section 2.4. Whilst the performance of VESPER relative to ERA5 is encouraging, one aspect of this is that VESPER has been trained directly on MODIS data whereas ERA5 has not. For this work we take MODIS data as our source of truth - as far as VESPER is concerned the MODIS data is reality, whereas of course the MODIS data has its own errors and systematics. The question of if a deep learning model could be used for forecasting is a very interesting one, but slightly beyond the scope of this study - we are primarily interested in quickly evaluating the accuracy of the fields that get passed to a dynamical model.
	\item \textbf{Section 3 Results}

	This section has been restructured to just contain the lake results, rather than the VESPER configuration as requested.

	A short discussion on how the  non-lake climate fields such as vegetation cover or orography are update in response to the update in the lake fields is not included at the end of Section 2.2.1 c.f. Aral sea. We have tried to condense this section, but generally there is lots to discuss and we do prefer to be thorough here. As suggested there is plenty of further discussion to be had on e.g. monthly lake maps, glaciers etc. but we defer this for a future study

	\item \textbf{Section 4 Discussion}

	We have added the referee's point about if VESPER and ECLand parametrisations would react similarly to changes in the input fields to the discussion. In short, this is a very interesting question. The use of a tool that was trained from ERA5 in model output of IFS requires the assumption that the statistical behaviour of the fields does not change from ERA to IFS as the ML model would otherwise be forced to extrapolate (which it will not be able to do). This is a fair assumption, but it would be interesting to investigate this quantitively in greater detail. We defer the investigation of this question to a future work and thank the referee for raising a good point.

\end{itemize}

### 2023-10-17 docs/submission/response_to_reviewer_comments/reviewer_1/response_reviewer_1.tex

\documentclass[10pt]{article}
\usepackage{vmargin}
\setpapersize{USletter}
\usepackage{ragged2e}

\author{Tom Kimpson \\ 
	\and Margarita Choulga\\ 
 \and Matthew Chantry\\
\and Gianpaolo Balsamo\\  
 \and Souhail Boussetta\\
  \and Peter Dueben\\
   \and Tim Palmer\\
}
\title{\normalsize Response to Reviewer  Comments
  concerning HESS submission egusphere-2022-1177}
\begin{document}
	\maketitle
\noindent Author response to referee reports for the paper egusphere-2022-1177,
entitled \textit{``Deep Learning for Verification of Earth-System Parametrisation of Water Bodies"}. \newline

\noindent We thank the referees for their reading, helpful criticism and suggestions towards the improvement of the manuscript. \newline

### 2023-10-17 docs/submission/template.tex
small first changes

\begin{abstract}
A purposely built deep learning algorithm for the Verification of Earth-System ParametERisation (VESPER) is used to assess recent upgrades of the global physiographic datasets underpinning the quality of the Integrated Forecasting System (IFS) of the European Centre for Medium-Range Weather Forecasts (ECMWF), which is used both in numerical weather prediction and climate reanalyses. A neural network regression model is trained to learn the mapping between the surface physiographic dataset plus the meteorology from ERA5, and the MODIS satellite skin temperature observations. Once trained, this tool is applied to rapidly assess the quality of upgrades of the land-surface scheme. Upgrades which improve the prediction accuracy of the machine learning tool indicate a reduction of the errors in the surface fields used as input to the surface parametrisation schemes. Conversely, incorrect specifications of the surface fields decrease the accuracy with which VESPER can make predictions.  We apply VESPER to assess the accuracy of recent upgrades of the permanent lake and glaciers covers as well as planned upgrades to represent seasonally varying water bodies (i.e. ephemeral lakes). We show that for grid-cells where the lake fields have been updated, the prediction accuracy in the land surface temperature (i.e mean absolute error difference between updated and original physiographic datasets) improves by 0.37 K on average, whilst for the subset of points where the lakes have been exchanged for bare ground (or vice versa) the improvement is 0.83 K. We also show that updates to the glacier cover improve the prediction accuracy by 0.22 K. We highlight how neural networks such as VESPER can assist the research and development of surface parametrizations and their input physiography to better represent Earth’s surface coupled processes in weather and climate models.

%Water bodies exchange mass and energy with the atmosphere and need to be accurately simulated in numerical weather prediction and climate modelling as they strongly influence the lower boundary conditions such as skin temperatures, and surface fluxes of heat and moisture near the surface. All the non-ocean water (resolved and sub-grid lakes and coastal waters) are represented in the Integrated Forecasting System (IFS) of the European Centre for Medium-Range Weather Forecasts (ECMWF) model, by the Fresh-water Lake (FLake) parametrisation, which treats $\sim$ 1/3 of the land. It is a continuous enterprise to update the surface parametrization schemes and their input fields to better represent small-scale processes. It is, however, difficult to quickly determine both the accuracy of an updated parametrisation, and the added value gained for the purposes of numerical modelling. In this work a neural network regression model is developed to learn the mapping between surface data from ERA5 and MODIS satellite observations of skin temperature. Once trained, the tool can be used to verify the quality of upgrades of the land-surface parametrisation scheme since large differences between the model and the machine learning tool indicate incorrect values, which are most often caused by errors in the surface fields that are used as input to the parametrisation scheme. We deploy this model to determine the accuracy of recent upgrades to the FLake parametrisation, namely the improved permanent lake cover and the capacity to represent seasonally varying water bodies (i.e. ephemeral lakes). We show that for grid-cells where the lake fields have been updated, the prediction accuracy in the land surface temperature improves by 0.37 K on average, whilst for the subset of points where the lakes have been exchanged for bare ground (or vice versa) the improvement is 0.84 K. We also show that updates to the glacier cover improve further the prediction accuracy by 0.22 K.

@@

%\copyrightstatement{TEXT} %% This section is optional and can be used for copyright transfers.

%\introduction  %% \introduction[modified heading if necessary]
\section{Introduction}
Accurate knowledge of the global surface physiography, including land, water and ice covers, and their characteristics, strongly determines the quality of surface and near-surface temperature simulations in weather and climate modelling. For instance, water bodies exchange mass and energy with the atmosphere and their thermal inertia strongly influence the lower boundary conditions such as skin temperatures, and surface fluxes of heat and moisture near the surface. Globally, there are $\sim$ 117 million lakes - defined as inland water bodies without lateral movement of water - making up around 3.7$\%$ of the Earth's land surface \citep{Verpoorter2014}. Their distribution is highly non-uniform, with the majority of lakes located between $45-75^{\circ}$N in the Boreal and Arctic regions. Lakes are highly important from the perspective of both numerical weather prediction and climate modelling as part of the EC-Earth model. For the latter, lakes generally influence the global carbon cycle as both sinks and sources of greenhouse gases; the majority of lakes are net heterotrophic (i.e. over saturated with carbon dioxide, CO$_2$), as a result of in lake respiration and so emit carbon into the atmosphere \citep{Pace2005,Tranvik2009}.  Total CO$_2$ emission from lakes is estimated at $1.25 - 2.30$ Pg of CO$_2$-equivalents annually \citep{DelSontro2018}, nearly $20 \%$ of global CO$_2$ fossil fuel emissions, whilst lakes account for 9-24 $\%$  of CH$_4$ emissions, the second largest natural source after wetlands \citep{Saunois2020}. These rates of greenhouse gas emission are expected to rise further if the eutrophication (i.e. nutrient concentration increase) of the Earth's lentic systems continues. With regards to weather, freezing and melting of the lake surface modifies the radiative and conductive properties and consequently affects the heat (latent, sensible) exchange and surface energy balance \citep{Franz2018,Huang2019,Peng2020}. Considering particular examples, over Lake Victoria convective activity is suppressed during the day and peaks at night, leading to intense, hazardous thunderstorms \citep{Thiery2015,Thiery_2017}; Lake Ladoga can generate low level clouds which can cause variability in the 2m temperature of up to 10 K \citep{Eerola2014}; the Laurentian Great Lakes can cause intense winter snow storms \citep{Notaro2013,Vavrus2013}. Moreover, as a result of the increased temperatures due to climate change, lakes become more numerous due to the melting of glaciers and permafrost. Additionally, the higher temperatures mean that previously permanent lake bodies become seasonal or intermittent. There is then evidently a huge potential return in the ability to accurately model the location, morphology and properties of lakes in weather and climate models. \newline

@@

\noindent Lakes are incorporated in Earth-system models via parametrisation. At ECMWF the  representation of lakes via parametrisation was first handled by introducing the Fresh water Lake model FLake \citep{Mironov2008} into the IFS. FLake treats all resolved inland waterbodies (i.e. lakes, reservoirs, rivers which are dominating in a grid-cell) and unresolved or sub-grid water (i.e. small inland waterbodies and sea/ocean coastal waters which are present but not dominating in a grid-cell). Its main drivers (input fields) are lake location and lake mean depth. The broad impact of the FLake model (i.e. areas where it is active) and the important role that waterbodies play in human life can be illustrated by analysing ECMWF fields of the fractional land sea mask and the inland waterbody cover alongside the population density field (i.e. inhabitants per km$^2$) based on the population count for 2015 from the Global Human Settlement Layers (GHSL), Population Grid 1975-2030 \citep{JRC100523,GHS} at 9 km horizontal resolution.

@@

\section{Constructing VESPER}\label{sec:2}
In order to rapidly assess the accuracy of new surface physiography fields and if their use in the model increase the accuracy with which we can make predictions, a neural network regression model (VESPER, hereafter) that can learn the mapping between a set of input features $x$ and targets $y$ is constructed. In this case the features are the atmospheric and surface model fields (such as 2 metre temperature from ERA5 reanalysis) and the surface physiographic fields (such as orography and vegetation cover used to produce ERA5 reanalysis). See Table \ref{table:definitions} for the full list of variables used. The target is the satellite land surface temperature (LST; skin temperature from MODIS Aqua Day MYD11A1 v006 collection). Once trained, VESPER can then make predictions about the skin temperature given a set of input variables (i.e. atmospheric and surface model fields, and surface physiographic fields). In turn, these predictions can then be compared against observations (i.e. satellite skin temperature) and VESPER's accuracy evaluated. By varying the number, type and values of the input features to VESPER and observing how the accuracy of its predictions change, some conclusions on if and how features can increase predictability of an actual atmospheric model can be drawn. Moreover, by isolating geographic regions where the predictions get worse with new/updated surface physiographic fields, areas where these fields might be erroneous or not informative enough can be identified. Due to the inherent stochasticity of training a neural network regression model it is also possible for different models to settle in different local minimums i.e. the network variance/noise. To understand the significance of this, every VESPER configuration was trained four times, each time with a different random seed. \newline

@@

\subsection{Features and targets}
VESPER's input feature selection (see Table \ref{table:definitions}) followed (i) permutation importance results for atmospheric and surface model fields - only fields with the highest importance were chosen; and (ii) expert choice for surface physiographic fields. As a first attempt it was decided to test the current methodology for lake related information, therefore fields that could be most affected by the presence or absence of water were selected, e.g. if lake had to be removed then some other surface had to appear (like bare ground, high or low vegetation, glacier or even ocean) and surface elevation had to change. Changes to the orographic fields will have important influences on temperature through e.g. wind, solar heating, etc. Lake depth changes are similarly important, influencing how a lake freezes, thaws, mixes and its overall dynamical range. VESPER's target selection followed globally available criteria and the satellite LST is quite well observed globally and with high temporal pattern (daily or even several times a day depending on the location).

@@

\subsubsection{Surface physiographic fields}\label{sec:surface_physio}
\noindent Surface physiographic fields have gridded information of the Earth’s surface properties (e.g. land-use, vegetation type and distribution) and represent surface heterogeneity in the ECLand of the IFS. They are used to compute surface turbulent fluxes (of heat, moisture and momentum) and skin temperature over different surfaces (vegetation, bare soil, snow, interception and water) and then to calculate an area-weighted average for the grid-box to couple with the atmosphere. To trigger all different parametrization schemes the ECMWF model uses a sets of physiographic fields, that do not depend on initial condition or forecast step. Most fields are constant; surface albedo is specified for 12 months to describe the seasonal cycle. Dependent on the origin, initial data comes at different resolutions and different projections, and is then first converted to a regular latitude-longitude grid (EPSG:4326) at $\sim$ 1km at Equator resolution, and secondly to a required grid and resolution. Surface physiographic fields used in this work consist of orographic, land, water, vegetation, soil, albedo fields and their difference between initial V15 and updated V20 field sets. See Tables \ref{table:definitions} and \ref{tab:datasources} for the full list of surface physiographic fields and their input sources; for more details see IFS documentation \citep{IFSdocs}. As this work is focused on assessing quality of inland water information, main surface physiographic fields are lake cover (derived from land-sea mask) and lake mean depth (see Table \ref{tab:datasources}).  \newline 
\begin{table*}
	\begin{tabularx}{\textwidth}{lXX}
		\toprule 
		Field category & V15 (initial) & V20 (updated) \\
		\hline
		Orographic & SRTM30 Shuttle Radar Topography Mission over 60°N-60°S; GLOBE: Global Land One-km Base Elevation Project data over 90-60°N; RAMP2: high-resolution Radarsat Antarctic Mapping Project Digital Elevation Model version 2 data \citep{Liu2015} over 60-90°S; BPRC: Byrd Polar Research Center over Greenland; IS 50V: Digital Map Database of Iceland over Iceland & As V15, with corrections of erroneous shift
		\\
		\hline 
			Land &	\textbf{glm:} GLCC: Global Land Cover Characteristics version 2.0 over 90°N-90°S except Iceland; Icelandic Meteorological Office (IMO) glacier mask 2013 over Iceland \newline 
		\textbf{lsm:} GlobCover2009 \citep{GLOBCOVER,arino2012glcm} over 85°N-60°S; RAMP2: high-resolution Radarsat Antarctic Mapping Project Digital Elevation Model version 2 data \citep{Liu2015} over 60-90°S; no land assumed over 90-85°N

		 &  \textbf{glm:} Norwegian Institute glacier data over Svalbard; Icelandic Meteorological Office (IMO) glacier mask 2017 over Iceland; GIMP: Greenland Ice Mapping Project data \citep{Howat2014} over Greenland; CryoSat-2 satellite glacier data \citep{Slater2018} over Antarctica (+ manual gap filling);  GLIMS: Global Land Ice Measurements from Space data \citep{glims} over rest of the globe  \newline 
		 \textbf{lsm:} GSWE: Global Surface Water Explorer \citep{GSWE}; glm \\
		\hline
				Water   & 
				\textbf{cl:} \textit{lsm} (ocean is separated at actual resolution by seeding and removing all connected grid-cells, includes the Caspian Sea, the Azov Sea, The American Great Lakes) \newline 
				\textbf{dl:} The Caspian Sea bathymetry; Global Relief Model ETOPO1 \citep{amante2009egrm} over the Great Lakes, the Azov Sea; GLDB: Global Lake DataBase version 1 \citep{Kourzeneva2012} over rest of the globe; 25 meters assumed over missing data grid-cells
		& \textbf{cl:} \textit{lsm} (ocean is separated at ~1km resolution by upgraded flooding algorithm following \cite{Choulga2019} \newline 
		\textbf{dl:} GEBCO: General Bathymetric Charts of the Ocean \citep{Weatherall2015} over the Caspian Sea and the Azov Sea; Global Relief Model ETOPO1 \citep{amante2009egrm} over the Great Lakes; GLDB: Global Lake DataBase version 3 \citep{Choulga2014} over rest of the globe; indirect estimates based on geological origin of lakes \citep{Choulga2014} over missing data grid-cells \\
		\hline 
		Vegetation & GLCC: Global Land Cover Characteristics version 1.2. Note that vegetation type represent only dominant type over grid-cell & As V15 \\ 
		\hline
		 Soil & DSMW: FAO/UNESCO Digital Soil Map of the world \citep{FAO}. Note that soil type represent only dominant type over grid-cell & As V15 \\ 
		 \hline 
		 Albedo & MODIS 5-year climatology \citep{SCHAAF2002135}; RossThickLiSparseReciprocal BRDF model. Note that Albedo values represent snow free surface albedo & As V15 \\ 
		\bottomrule
	\end{tabularx}
	\caption{List of input datasets for the surface physiographic fields for V15 and V20 field sets. V15X and V20X are identical to V15 and V20 respectively, but with the addition of saline lake cover, and monthly varying lake cover fields. \newline }
	\label{tab:datasources}
\end{table*}

@@

To expand V15 and V20 lake description (to V15X and V20X respectively) their salinity and time variability information was generated. Even though static permanent water fits better to describe inland water distribution on average all year round, some areas (in Tropics especially) could  benefit from having monthly varying information as they have a very strong seasonal cycle, when size, shape and depth of a lake changes over the course of the year, leading to a significant change in modelling the lake temperature response. Similarly, saline lakes behave very differently to fresh water lakes since increased salt concentrations affect the density, specific heat capacity, thermal conductivity, and turbidity, as well as evaporation rates, ice formation and ultimately the surface temperature. These two properties of time variability and salinity are often related; it is common for saline lakes to fill and dry out over the course of the season, which naturally also affects the relative saline concentration of the lake itself. To create a monthly varying lake cover first 12 monthly fractional land-sea masks based on JRC Monthly Water History v1.3 maps for 2010-2020 were created. Since the annual lake maps were created taking into account a lot of additional sources the extra condition on the monthly maps that the monthly water is equal or greater than permanent water distribution from fractional land-sea mask is enforced. To create an inland salt lake cover map, the GLDBv3 salt lake list was used. First, in order to identify separate lakes on $\sim$ 1km resolution lake cover (by ``lake cover" we refer the maximum lake distribution based on 12 monthly-varying lake covers), small sub-grid lakes and large lake coasts are masked, i.e. grid-cells that have water fraction less than 0.25. Next, number of connected grid-cells in each lake (i.e. connected with sides only) is computed. Then only lakes that have 100 and more connected grid-cells are vectorised, as at ERA5 resolution of ∼31km the grid-cells are quite large and can include a mixture of freshwater and saline lakes. Finally, saline lake vectors are selected by filtering vectors which have no saline lake point from GLDBv3 located – in total 147 large salt lake vectors, which were further used to filter non-saline lakes at ~1km resolution lake cover, finally aggregated to ~31km resolution. In the future it is planned to revisit this field and extend the list to include additional data. Note that all non-lake related climate fields such as vegetation cover or orography were updated in V20 field set compared to V15 only in relation to the changing lake fields (i.e. if fraction of lake in the grid cell increased then other fractions like vegetation or bare ground should have increased accordingly).

>>>>

\begin{abstract}
A purposely built deep learning algorithm for the Verification of Earth-System ParametERisation (VESPER) is used to assess recent upgrades of the global physiographic datasets underpinning the quality of the Integrated Forecasting System (IFS) of the European Centre for Medium-Range Weather Forecasts (ECMWF), which is used both in numerical weather prediction and climate reanalyses. A neural network regression model is trained to learn the mapping between the surface physiographic dataset plus the meteorology from ERA5, and the MODIS satellite skin temperature observations. Once trained, this tool is applied to rapidly assess the quality of upgrades of the land-surface scheme. Upgrades which improve the prediction accuracy of the machine learning tool indicate a reduction of the errors in the surface fields used as input to the surface parametrisation schemes. Conversely, incorrect specifications of the surface fields decrease the accuracy with which VESPER can make predictions.  We apply VESPER to assess the accuracy of recent upgrades of the permanent lake and glaciers covers as well as planned upgrades to represent seasonally varying water bodies (i.e. ephemeral lakes). We show that for grid-cells where the lake fields have been updated, the prediction accuracy in the land surface temperature (as quantified by the mean absolute error) improves by 0.37 K on average, whilst for the subset of points where the lakes have been completely removed and replaced with bare ground the improvement is 0.83 K. We also show that updates to the glacier cover improve the prediction accuracy by 0.22 K. We highlight how neural networks such as VESPER can assist the research and development of surface parametrizations and their input physiography to better represent Earth’s surface coupled processes in weather and climate models.

@@

\section{Introduction}
Accurate knowledge of the global surface physiography, including land, water and ice covers, and their characteristics, strongly determines the quality of surface and near-surface temperature simulations in weather and climate modelling. For instance, water bodies exchange mass and energy with the atmosphere and their thermal inertia strongly influences the lower boundary conditions such as skin temperatures, and surface fluxes of heat and moisture near the surface. Globally, there are $\sim$ 117 million lakes - defined as inland water bodies without lateral movement of water - making up around 3.7$\%$ of the Earth's land surface \citep{Verpoorter2014}. Their distribution is highly non-uniform, with the majority of lakes located between $45-75^{\circ}$N in the Boreal and Arctic regions. Lakes are highly important from the perspective of both numerical weather prediction and climate modelling as part of the EC-Earth model. For the latter, lakes generally influence the global carbon cycle as both sinks and sources of greenhouse gases; the majority of lakes are net heterotrophic (i.e. oversaturated with carbon dioxide, CO$_2$), as a result of in lake respiration and so emit carbon into the atmosphere \citep{Pace2005,Tranvik2009}.  Total CO$_2$ emission from lakes is estimated at $1.25 - 2.30$ Pg of CO$_2$-equivalents annually \citep{DelSontro2018}, nearly $20 \%$ of global CO$_2$ fossil fuel emissions, whilst lakes account for 9-24 $\%$  of CH$_4$ emissions, the second largest natural source after wetlands \citep{Saunois2020}. These rates of greenhouse gas emission are expected to rise further if the eutrophication (i.e. nutrient concentration increase) of the Earth's lentic systems continues. With regards to weather, freezing and melting of the lake surface modifies the radiative and conductive properties and consequently affects the heat (latent, sensible) exchange and surface energy balance \citep{Franz2018,Huang2019,Peng2020}. Considering particular examples, over Lake Victoria convective activity is suppressed during the day and peaks at night, leading to intense, hazardous thunderstorms \citep{Thiery2015,Thiery_2017}; Lake Ladoga can generate low level clouds which can cause variability in the 2m temperature of up to 10 K \citep{Eerola2014}; the Laurentian Great Lakes can cause intense winter snow storms \citep{Notaro2013,Vavrus2013}. Moreover, as a result of the increased temperatures due to climate change, lakes become more numerous due to the melting of glaciers and permafrost. Additionally, the higher temperatures mean that previously permanent lake bodies become seasonal or intermittent. There is then evidently a huge potential return in the ability to accurately model the location, morphology and properties of lakes in weather and climate models. \newline

@@

\noindent Lakes are incorporated in Earth-system models via parametrisation. At ECMWF the  representation of lakes via parametrisation was first handled by introducing the Fresh water Lake model FLake \citep{Mironov2008} into the IFS. FLake treats all resolved inland waterbodies (i.e. lakes, reservoirs, rivers which are dominating in a grid-cell) and unresolved or sub-grid water (i.e. small inland waterbodies and sea/ocean coastal waters which are present but not dominating in a grid-cell). Its main input fields are lake location and lake mean depth. The broad impact of the FLake model (i.e. areas where it is active) and the important role that waterbodies play in human life can be illustrated by analysing ECMWF fields of the fractional land sea mask and the inland waterbody cover alongside the population density field (i.e. inhabitants per km$^2$) based on the population count for 2015 from the Global Human Settlement Layers (GHSL), Population Grid 1975-2030 \citep{JRC100523,GHS} at 9 km horizontal resolution.

@@

\section{Constructing VESPER}\label{sec:2}
In order to rapidly assess the accuracy of new surface physiography fields and if their use in the model increase the accuracy with which we can make predictions, a neural network regression model (VESPER, hereafter) that can learn the mapping between a set of input features $x$ and targets $y$ is constructed. In this case the features are the atmospheric and surface model fields, such as 2 metre temperature from ERA5 reanalysis, and the surface physiographic fields, such as orography and vegetation cover used to produce ERA5 reanalysis. See Table \ref{table:definitions} for the full list of variables used. The target is the satellite land surface temperature (LST; skin temperature from MODIS Aqua Day MYD11A1 v006 collection). Once trained, VESPER can then make predictions about the skin temperature given a set of input variables, i.e. atmospheric and surface model fields, and surface physiographic fields. In turn, these predictions can then be compared against observations, i.e. satellite skin temperature, and VESPER's accuracy evaluated. By varying the number, type and values of the input features to VESPER and observing how the accuracy of its predictions change, some conclusions on if and how features can increase predictability of an actual atmospheric model can be drawn. Moreover, by isolating geographic regions where the predictions get worse with new/updated surface physiographic fields, areas where these fields might be erroneous or not informative enough can be identified. Due to the inherent stochasticity of training a neural network regression model it is also possible for different models to settle in different local minimums i.e. the network variance/noise. To understand the significance of this, every VESPER configuration was trained four times, each time with a different random seed. \newline

@@

\subsection{Features and targets}
VESPER's input feature selection (see Table \ref{table:definitions}) followed (i) permutation importance results for atmospheric and surface model fields - only fields with the highest importance were chosen; and (ii) expert choice for surface physiographic fields. As a first attempt it was decided to test the current methodology for lake related information, therefore fields that could be most affected by the presence or absence of water were selected, e.g. if lake had to be removed then some other surface had to appear, such as bare ground, high or low vegetation, glacier or even ocean. Moreover the surface elevation had to change. Changes to the orographic fields will have important influences on temperature through e.g. wind, solar heating, etc. Lake depth changes are similarly important, influencing how a lake freezes, thaws, mixes and its overall dynamical range. VESPER's target selection followed globally available criteria and the satellite LST is quite well observed globally and with high temporal pattern, daily or even several times a day depending on the location.

@@

\subsubsection{Surface physiographic fields}\label{sec:surface_physio}
\noindent Surface physiographic fields have gridded information of the Earth’s surface properties (e.g. land-use, vegetation type and distribution) and represent surface heterogeneity in the ECLand of the IFS. They are used to compute surface turbulent fluxes (of heat, moisture and momentum) and skin temperature over different surfaces (vegetation, bare soil, snow, interception and water) and then to calculate an area-weighted average for the grid-box to couple with the atmosphere. To trigger all different parametrization schemes the ECMWF model uses a sets of physiographic fields, that do not depend on initial condition of each forecast run, or the forecast step. Most fields are constant; surface albedo is specified for 12 months to describe the seasonal cycle. Dependent on the origin, initial data comes at different resolutions and different projections, and is then first converted to a regular latitude-longitude grid (EPSG:4326) at $\sim$ 1km at Equator resolution, and secondly to a required grid and resolution. Surface physiographic fields used in this work consist of orographic, land, water, vegetation, soil, albedo fields and their difference between initial V15 and updated V20 field sets. See Tables \ref{table:definitions} and \ref{tab:datasources} for the full list of surface physiographic fields and their input sources; for more details see IFS documentation \citep{IFSdocs}. As this work is focused on assessing quality of inland water information, main surface physiographic fields are lake cover (derived from land-sea mask) and lake mean depth (see Table \ref{tab:datasources}).  \newline 
\begin{table*}
	\begin{tabularx}{\textwidth}{lXX}
		\toprule 
		Field category & V15 (initial) & V20 (updated) \\
		\hline
		Orographic & SRTM30 Shuttle Radar Topography Mission over 60°N-60°S; GLOBE: Global Land One-km Base Elevation Project data over 90-60°N; RAMP2: high-resolution Radarsat Antarctic Mapping Project Digital Elevation Model version 2 data \citep{Liu2015} over 60-90°S; BPRC: Byrd Polar Research Center over Greenland; IS 50V: Digital Map Database of Iceland over Iceland & As V15, with corrections of erroneous shift
		\\
		\hline 
			Land &	\textbf{glm:} GLCC: Global Land Cover Characteristics version 2.0 over 90°N-90°S except Iceland; Icelandic Meteorological Office (IMO) glacier mask 2013 over Iceland \newline 
		\textbf{lsm:} GlobCover2009 \citep{GLOBCOVER,arino2012glcm} over 85°N-60°S; RAMP2: high-resolution Radarsat Antarctic Mapping Project Digital Elevation Model version 2 data \citep{Liu2015} over 60-90°S; no land assumed over 90-85°N

		 &  \textbf{glm:} Norwegian Institute glacier data over Svalbard; Icelandic Meteorological Office (IMO) glacier mask 2017 over Iceland; GIMP: Greenland Ice Mapping Project data \citep{Howat2014} over Greenland; CryoSat-2 satellite glacier data \citep{Slater2018} over Antarctica (+ manual gap filling);  GLIMS: Global Land Ice Measurements from Space data \citep{glims} over rest of the globe  \newline 
		 \textbf{lsm:} GSWE: Global Surface Water Explorer \citep{GSWE}; glm \\
		\hline
				Water   & 
				\textbf{cl:} \textit{lsm} (ocean is separated at actual resolution by seeding and removing all connected grid-cells, includes the Caspian Sea, the Azov Sea, The American Great Lakes) \newline 
				\textbf{dl:} The Caspian Sea bathymetry; Global Relief Model ETOPO1 \citep{amante2009egrm} over the Great Lakes, the Azov Sea; GLDB: Global Lake DataBase version 1 \citep{Kourzeneva2012} over rest of the globe; 25 meters assumed over missing data grid-cells
		& \textbf{cl:} \textit{lsm} (ocean is separated at ~1km resolution by upgraded flooding algorithm following \cite{Choulga2019} \newline 
		\textbf{dl:} GEBCO: General Bathymetric Charts of the Ocean \citep{Weatherall2015} over the Caspian Sea and the Azov Sea; Global Relief Model ETOPO1 \citep{amante2009egrm} over the Great Lakes; GLDB: Global Lake DataBase version 3 \citep{Choulga2014} over rest of the globe; indirect estimates based on geological origin of lakes \citep{Choulga2014} over missing data grid-cells \\
		\hline 
		Vegetation & GLCC: Global Land Cover Characteristics version 1.2. Note that vegetation type represent only dominant type over grid-cell & As V15 \\ 
		\hline
		 Soil & DSMW: FAO/UNESCO Digital Soil Map of the world \citep{FAO}. Note that soil type represent only dominant type over grid-cell & As V15 \\ 
		 \hline 
		 Albedo & MODIS 5-year climatology \citep{SCHAAF2002135}; RossThickLiSparseReciprocal BRDF model. Note that Albedo values represent snow free surface albedo & As V15 \\ 
		\bottomrule
	\end{tabularx}
	\caption{List of input datasets for the surface physiographic fields for V15 and V20 field sets. V15X and V20X are identical to V15 and V20 respectively, but with the addition of saline lake cover, and monthly varying lake cover fields. \newline }
	\label{tab:datasources}
\end{table*}

@@

To expand V15 and V20 lake description (to V15X and V20X respectively) their salinity and time variability information was generated. Even though static permanent water fits better to describe inland water distribution on average all year round, some areas (in Tropics especially) could  benefit from having monthly varying information as they have a very strong seasonal cycle, when size, shape and depth of a lake changes over the course of the year, leading to a significant change in modelling the lake temperature response. Similarly, saline lakes behave very differently to fresh water lakes since increased salt concentrations affect the density, specific heat capacity, thermal conductivity, and turbidity, as well as evaporation rates, ice formation and ultimately the surface temperature. These two properties of time variability and salinity are often related; it is common for saline lakes to fill and dry out over the course of the season, which naturally also affects the relative saline concentration of the lake itself. To create a monthly varying lake cover first 12 monthly fractional land-sea masks based on JRC Monthly Water History v1.3 maps for 2010-2020 were created. Since the annual lake maps were created taking into account a lot of additional sources the extra condition on the monthly maps that the monthly water is equal or greater than permanent water distribution from fractional land-sea mask is enforced. To create an inland salt lake cover map, the GLDBv3 salt lake list was used. First, in order to identify separate lakes on $\sim$ 1km resolution lake cover (by ``lake cover" we refer the maximum lake distribution based on 12 monthly-varying lake covers), small sub-grid lakes and large lake coasts are masked, i.e. grid-cells that have water fraction less than 0.25. Next, the number of connected grid-cells in each lake (i.e. connected with sides only) is computed. Then only lakes that have 100 and more connected grid-cells are vectorised, as at ERA5 resolution of ∼31km the grid-cells are quite large and can include a mixture of freshwater and saline lakes. Finally, saline lake vectors are selected by filtering vectors which have no saline lake point from GLDBv3 located – in total 147 large salt lake vectors, which were further used to filter non-saline lakes at ~1km resolution lake cover, finally aggregated to ~31km resolution. In the future it is planned to revisit this field and extend the list to include additional data. Note that all non-lake related climate fields such as vegetation cover or orography were updated in V20 field set compared to V15 only in relation to the changing lake fields (i.e. if fraction of lake in the grid cell increased then other fractions like vegetation or bare ground should have decreased accordingly).
